Continuous gravitational waves in the lab: recovering audio signals with a table-top~optical microphone

### 2021-09-23 paper/ifo-complexAudio.tex
moving citations to after puctuation

We use the Michelson interferometer as an ``optical microphone'', where a laser is used to detect sound instead of the electrical components in a conventional microphone.
The only change to our apparatus is replacing the webcam with a photodiode to allow us to capture higher frequencies necessary for speech and music (see Section~\ref{sec:photodiode}). 
Optical microphones have precedence in the laser microphones~\cite{laser_microphone} which are (or were historically) used in the defense industry and operate on a variety of related principles. 
%We use the term ``optical microphone'' to distinguish this technology from those previous technologies.
Our objective is to play a recording of speech or music through the speaker attached to mirror M2 (see Fig.~\ref{fig:ifo_schematic_webcam}), record the resulting interference pattern, and then recover the original signal via a selection of signal processing techniques.

%This analysis is indirectly related to the advanced signal processing techniques required to extract gravitational-wave signals. % citation?
The apparatus serves as an independent demonstration for a broader physics and engineering audience, particularly in undergraduate laboratories. 
We describe additional hardware components required for this demonstration in Section~\ref{sec:photodiode} and the initial results in Section~\ref{sec:initialResultsOpMic}. 
We consider a selection of filter techniques, details of which along with a summary of digital signal processing resources can be found in the Supplementary Material.
 % We also refer the reader to a selection of digital signal processing resources for further information. 
In Section~\ref{sec:opticalMicResults}, we present the two best-performing techniques from the Supplementary Material.

@@

We place an OSRAM BPW21 photodiode in reverse-bias over an LM358 op-amp which together produce a voltage that depends on the incident intensity. 
The photodiode records the interference pattern at roughly the same off-center position as the webcam in Sections~\ref{sec:single_tone} and~\ref{sec:viterbi_wandering}, again chosen arbitrarily.
% I think this is probably not needed as I believe the editor comment is referring to the cloth screen over the photodiode, which is addressed below. 
%The photodiode is mounted on a piece of cloth stretched over a plastic frame, with the electrical leads connected underneath the cloth.
%This cloth screen mount was the ``grill cloth'' from the front of a dismantled commercial speaker, re-purposed. 
The voltage signal from the photo-detector is captured by an MCP3008 $10$-bit analog-to-digital converter (ADC) connected to a Raspberry Pi Model 3 v1.2, which provides a convenient means to record the photodiode data.
Together, the circuit samples the signal at $\sim 16\,{\rm kHz}$. 
Resources for using the Raspberry~Pi and photodiode, including a circuit diagram, are described in the Supplementary Material.

Sampling any frequency component of the analog signal above the Nyquist frequency of $8\,{\rm kHz}$ leads to aliasing (folding of frequencies greater than half the sampling rate) into the detected range. We include an anti-aliasing Sallen-Key filter tuned to $16\,{\rm kHz}$ before the ADC to prevent this from happening.
This component attenuates any frequencies above $8\,{\rm kHz}$ before they are digitally sampled. We also place a cloth screen over the face of the photodiode to reduce the incident intensity and avoid saturating the ADC -- an improvised, physical solution that could instead be replaced by scaling down the voltage electronically. This cloth screen was re-purposed grill cloth from a commercial speaker.
% It’s not clear what limited the sample rate to 16kHz, but it was likely non-optimal reading of the ADC by the Pi script.
% The ADC used (MCP3008) is quoted at 200kHz (or ksps, kilo samples per second).

%An updated schematic and a photograph of the entire optical microphone are shown in the left and right panels of Fig.~\ref{fig:ifo_schematic_podo}, respectively.
%See Appendix~\ref{app:circuit_diagram} for a diagram of the circuit.

@@

%The source audio was played through the speaker with the optical microphone recording for the full duration of the audio (around a minute long for most sources). 
We test the optical microphone with a variety of recordings, including the speech of different people and music ranging from simple melodies and rhythms to songs. 
During recordings, care is taken to minimize activity around the demonstration to reduce environmental noise coupling into the interferometer. 
The timeseries data is then directly converted to a .wav file and played as an audio recording using the \texttt{scipy.io.wavfile.write} function in Python (see Appendix~\ref{app:code}). % moved refs \cite{scipy,python}. 
When processing the results, we restrict our analysis to only the first $10\,{\rm s}$ of each observation (for efficiency), and only plot the first second here.

@@

\begin{figure}
	\includegraphics[width=.5\textwidth]{figures/psd_podo_14_6.pdf}
	\caption{\label{fig:psd_noise}
Power spectral density (PSD) of background noise from the optical microphone (with the speaker off). 
% Bottom: the PSD after applying a Butterworth bandpass filter (bottom panel) between the two frequencies marked with red, dashed lines. 
We see strong power from the $50\,{\rm Hz}$ mains hum and its harmonics (most likely from the photodiode circuit and the room’s lighting and cooling). Otherwise, the PSD is fairly white except for a peak at around 0.75~kHz. 
% After filtering we see strong attenuation (at least 3~dB) of all frequencies outside the band, but little change to the harmonics within the band.
}
\end{figure}

@@

% https://en.wikipedia.org/wiki/Filter_(signal_processing)
In signal processing, the ideal filter would be one that:
(i) completely attenuates the undesired parts of the spectrum, 
(ii) does not change the rest of the spectrum, and 
(iii) smoothly transitions between these regions, as to not damage the time domain signal when seen under convolution. 
However, these three conditions cannot all hold at once. For example, if conditions (i) and (ii) hold, then the filter must be discontinuous at the boundary of the undesired region but this implies that the filter has ``infinite latency'' and so will affect (or damage) the time domain signal for infinite time~\cite{10.5555/151045}. Therefore, any filter must compromise between these three conditions. For speech intelligibility, this means that either: (i) some noise remains in the filtered recording, (ii) some of the speech content is lost as certain important frequencies are attenuated, or (iii) the speech is somewhat distorted in time.
All three of these cases can, when taken to the extreme, make the speech intelligibility worse than the unfiltered recording. 
Therefore, we choose filters that compromise between achieving the three conditions.
% All three of these cases can, in extreme cases, worsen speech intelligibility beyond the unfiltered recording and so we choose filters that compromise between achieving the three conditions.

>>>>

We use the Michelson interferometer as an ``optical microphone'', where a laser is used to detect sound instead of the electrical components in a conventional microphone.
The only change to our apparatus is replacing the webcam with a photodiode to allow us to capture higher frequencies necessary for speech and music (see Section~\ref{sec:photodiode}). 
Optical microphones have precedence in the laser microphones~\cite{laser_microphone} which are (or were historically) used in the defense industry and operate on a variety of related principles. 
Our objective is to play a recording of speech or music through the speaker attached to mirror M2 (see Fig.~\ref{fig:ifo_schematic_webcam}), record the resulting interference pattern, and then recover the original signal via a selection of signal processing techniques.

The apparatus serves as an independent demonstration for a broader physics and engineering audience, particularly in undergraduate laboratories. 
We describe additional hardware components required for this demonstration in Section~\ref{sec:photodiode} and the initial results in Section~\ref{sec:initialResultsOpMic}. 
We consider a selection of filter techniques, details of which along with a summary of digital signal processing resources can be found in the Supplementary Material. 
In Section~\ref{sec:opticalMicResults}, we present the two best-performing techniques from the Supplementary Material.

@@

We place an OSRAM BPW21 photodiode in reverse-bias over an LM358 op-amp which together produce a voltage that depends on the incident intensity. 
The photodiode records the interference pattern at roughly the same off-center position as the webcam in Sections~\ref{sec:single_tone} and~\ref{sec:viterbi_wandering}, again chosen arbitrarily. 
The voltage signal from the photo-detector is captured by an MCP3008 $10$-bit analog-to-digital converter (ADC) connected to a Raspberry Pi Model 3 v1.2, which provides a convenient means to record the photodiode data.
Together, the circuit samples the signal at $\sim 16\,{\rm kHz}$. 
Resources for using the Raspberry~Pi and photodiode, including a circuit diagram, are described in the Supplementary Material.

Sampling any frequency component of the analog signal above the Nyquist frequency of $8\,{\rm kHz}$ leads to aliasing (folding of frequencies greater than half the sampling rate) into the detected range. We include an anti-aliasing Sallen-Key filter tuned to $16\,{\rm kHz}$ before the ADC to prevent this from happening.
This component attenuates any frequencies above $8\,{\rm kHz}$ before they are digitally sampled. We also place a cloth screen over the face of the photodiode to reduce the incident intensity and avoid saturating the ADC -- an improvised, physical solution that could instead be replaced by scaling down the voltage electronically. This cloth screen was re-purposed grill cloth from a commercial speaker.

@@

We test the optical microphone with a variety of recordings, including the speech of different people and music ranging from simple melodies and rhythms to songs. 
During recordings, care is taken to minimize activity around the demonstration to reduce environmental noise coupling into the interferometer. 
The timeseries data is then directly converted to a .wav file and played as an audio recording using the \texttt{scipy.io.wavfile.write} function in Python (see Appendix~\ref{app:code}).
When processing the results, we restrict our analysis to only the first $10\,{\rm s}$ of each observation (for efficiency), and only plot the first second here.

@@

\begin{figure}
	\includegraphics[width=.5\textwidth]{figures/psd_podo_14_6.pdf}
	\caption{\label{fig:psd_noise}
Power spectral density (PSD) of background noise from the optical microphone (with the speaker off). 
We see strong power from the $50\,{\rm Hz}$ mains hum and its harmonics (most likely from the photodiode circuit and the room’s lighting and cooling). Otherwise, the PSD is fairly white except for a peak at around 0.75~kHz.
}
\end{figure}

@@

In signal processing, the ideal filter would be one that:
(i) completely attenuates the undesired parts of the spectrum, 
(ii) does not change the rest of the spectrum, and 
(iii) smoothly transitions between these regions, as to not damage the time domain signal when seen under convolution. 
However, these three conditions cannot all hold at once. 
For example, if conditions (i) and (ii) hold, then the filter must be discontinuous at the boundary of the undesired region but this implies that the filter has ``infinite latency'' and so will affect (or damage) the time domain signal for infinite time.\cite{10.5555/151045}
Therefore, any filter must compromise between these three conditions. 
For speech intelligibility, this means that either: (i) some noise remains in the filtered recording, (ii) some of the speech content is lost as certain important frequencies are attenuated, or (iii) the speech is somewhat distorted in time.
All three of these cases can, when taken to the extreme, make the speech intelligibility worse than the unfiltered recording. 
Therefore, we choose filters that compromise between achieving the three conditions.

### 2021-09-23 paper/ifo-conclusions.tex
moving citations to after puctuation

In this paper, we use a table-top Michelson interferometer as an analog to a gravitational-wave detector, demonstrating signal processing techniques used within the gravitational-wave community.
We explore the use of the interferometer as an optical microphone and consider a more general treatment of signal processing with complex audio signals, which can also serve as a distant analog to un-modeled gravitational-wave burst signals, e.g., from supernovae.
% justify ``un-modeled''? - james
The demonstration can be adapted for use in both the physics and engineering undergraduate laboratory, providing opportunities for cross-disciplinary teaching. 
Additionally, it can be used as a tool for explaining gravitational-wave research to a wider, non-specialist audience.

\begin{comment}

The demonstration presented has the potential to be adapted for use in the undergraduate laboratory.
It can be used to teach topics of interest to physics and electrical engineering students including interferometer physics; gravitational-wave detection, searches, and analysis; and signal processing with filters and speech enhancement techniques. 
The increased excitement and public interest in the field of gravitational-wave research in recent years mean that this demonstration may also be adapted as a tool for explaining gravitational-wave research to a wider, non-specialist audience.

\end{comment}

>>>>

In this paper, we use a table-top Michelson interferometer as an analog to a gravitational-wave detector, demonstrating signal processing techniques used within the gravitational-wave community.
We explore the use of the interferometer as an optical microphone and consider a more general treatment of signal processing with complex audio signals, which can also serve as a distant analog to minimally gravitational-wave burst signals, e.g., from supernovae.
% justify ``un-modeled''? - james
% fair point, it's the lingo used in LIGO, but really it's more accurate to say minimally modelled - Han
The demonstration can be adapted for use in both the physics and engineering undergraduate laboratory, providing opportunities for cross-disciplinary teaching. 
Additionally, it can be used as a tool for explaining gravitational-wave research to a wider, non-specialist audience.

### 2021-09-23 paper/ifo-futureWork.tex
moving citations to after puctuation

\begin{figure*}
    \includegraphics[width=.8\textwidth]{figures/pipeline_nobox.pdf}
	%\includegraphics[height=.8\textwidth, angle=-90]{figures/pipeline.pdf}
	\caption{\label{fig:pipeline_highlighted}
Flowchart for the signal through the optical microphone. 
A transfer function of the system would have to account for each stage and exploration of this is an aim of future work. 
Not shown are the signal processing filters that are performed on the PC once the data is transferred off the Raspberry~Pi.
}
\end{figure*}

@@

Isolating the experiment from interference may provide some noise reduction. 
The Raspberry Pi could be powered with a commercial battery instead of mains power, which may help to isolate the circuit from the mains as all other components draw power from the Raspberry~Pi (the laser is DC-powered through a converter).
Transferring the circuit from a breadboard to a printed circuit board can also reduce interference~\cite{elfekey2013design}.
A feedback loop control system can also be used to suppress any unwanted movement of the interferometer components~\citep{abbott2017exploring, Sekiguchi:2016bmv, verhoeven2009robust}, similar to methods used to isolate gravitational-wave observatories.

@@

A voluminous literature exists on using hidden Markov models trained on phonemes to recognize speech~\cite{HMM_english}. 
These could perform the final stage of speech enhancement for the optical microphone. 
This coincidentally connects back to the Viterbi algorithm in Section~\ref{sec:viterbi_wandering}, which is also underpinned by a hidden Markov model, albeit of a different type. Alternatively, machine learning solutions exist throughout the field that can compete with statistical techniques such as the logMMSE estimator~\cite{SEGAN}.

@@

Returning to the demonstration of gravitational wave analysis, single interferometers cannot detect directional information for signals; a large proportion of the directional information in gravitational-wave detections comes from the time delay offset between the signal being recorded at two or more detectors~\cite{GW150914}.
We could therefore extend this analysis to include data from two interferometers to extract directional information.
This would require increasing the sensitivity of the optical microphone to pick up the signal from a distant source instead of from a speaker attached directly to one of the mirrors of the interferometer.
Another extension would be to demonstrate the Doppler effect of the Earth's motion around the Sun, which needs to be considered in continuous-wave searches (see Section~\ref{sec:realCWSearches} and Ref.~\cite{JKS:1998}). 
One approach could be to modify the input audio signal to simulate Doppler modulation. 
% did we cut the turn-table idea? that was fun! - james

>>>>

\begin{figure*}
    \includegraphics[width=.8\textwidth]{figures/pipeline_nobox.pdf}
	\caption{\label{fig:pipeline_highlighted}
Flowchart for the signal through the optical microphone. 
A transfer function of the system would have to account for each stage and exploration of this is an aim of future work. 
Not shown are the signal processing filters that are performed on the PC once the data is transferred off the Raspberry~Pi.
}
\end{figure*}

@@

Isolating the experiment from interference may provide some noise reduction. 
The Raspberry Pi could be powered with a commercial battery instead of mains power, which may help to isolate the circuit from the mains as all other components draw power from the Raspberry~Pi (the laser is DC-powered through a converter).
Transferring the circuit from a breadboard to a printed circuit board can also reduce interference.\cite{elfekey2013design}
A feedback loop control system can also be used to suppress any unwanted movement of the interferometer components,\citep{abbott2017exploring, Sekiguchi:2016bmv, verhoeven2009robust} similar to methods used to isolate gravitational-wave observatories.

@@

A voluminous literature exists on using hidden Markov models trained on phonemes to recognize speech.\cite{HMM_english}
These could perform the final stage of speech enhancement for the optical microphone. 
This coincidentally connects back to the Viterbi algorithm in Section~\ref{sec:viterbi_wandering}, which is also underpinned by a hidden Markov model, albeit of a different type. Alternatively, machine learning solutions exist throughout the field that can compete with statistical techniques such as the logMMSE estimator.\cite{SEGAN}

@@

Returning to the demonstration of gravitational wave analysis, single interferometers cannot detect directional information for signals; a large proportion of the directional information in gravitational-wave detections comes from the time delay offset between the signal being recorded at two or more detectors.\cite{GW150914}
We could therefore extend this analysis to include data from two interferometers to extract directional information.
This would require increasing the sensitivity of the optical microphone to pick up the signal from a distant source instead of from a speaker attached directly to one of the mirrors of the interferometer.
Another extension would be to demonstrate the Doppler effect of the Earth's motion around the Sun, which needs to be considered in continuous-wave searches (see Section~\ref{sec:realCWSearches} and Ref.~\cite{JKS:1998}). 
One approach could be to modify the input audio signal to simulate Doppler modulation. 
% did we cut the turn-table idea? that was fun! - james
% yeah afraid so - it was nice, but the editor did not like it! :( - Hannah

### 2021-09-23 paper/ifo-singleTone.tex
moving citations to after puctuation

Continuous-wave searches look for nearly monochromatic signals~\cite{JKS:1998}. In this section, we consider a simple sinusoidal tone at a single, constant frequency.

>>>>

Continuous-wave searches look for nearly monochromatic signals.\cite{JKS:1998} 
In this section, we consider a simple sinusoidal tone at a single, constant frequency.

### 2021-09-23 paper/ifo-tableTopGWs.tex
moving citations to after puctuation

Gravitational-wave detectors such as LIGO and Virgo are large, complex experiments. 
However, their design is fundamentally based on the Michelson interferometer, an optical configuration commonly used in undergraduate laboratories. 
In a Michelson interferometer, laser light is split by a beamsplitter into two perpendicular arms, shown in Fig.~\ref{fig:ifo_schematic_webcam}. 
Mirrors at the end of each arm reflect the two beams back to the beamsplitter where they recombine to produce an interference pattern.
The resulting interference pattern is dependent on the relative distance traveled by the beams. 
Current generation gravitational-wave observatories have kilometer-scale arms (with arm lengths $4\,{\rm km}$ and $3\,{\rm km}$ at LIGO and Virgo, respectively). 
They can measure minuscule changes in distance due to gravitational waves; e.g., the first detection of a binary black hole merger produced a strain of $10^{-21}$~\cite{GW150914}, corresponding to a change in distance equal to a fraction of the width of a proton.

>>>>

Gravitational-wave detectors such as LIGO and Virgo are large, complex experiments. 
However, their design is fundamentally based on the Michelson interferometer, an optical configuration commonly used in undergraduate laboratories. 
In a Michelson interferometer, laser light is split by a beamsplitter into two perpendicular arms, shown in Fig.~\ref{fig:ifo_schematic_webcam}. 
Mirrors at the end of each arm reflect the two beams back to the beamsplitter where they recombine to produce an interference pattern.
The resulting interference pattern is dependent on the relative distance traveled by the beams. 
Current generation gravitational-wave observatories have kilometer-scale arms (with arm lengths $4\,{\rm km}$ and $3\,{\rm km}$ at LIGO and Virgo, respectively). 
They can measure minuscule changes in distance due to gravitational waves; e.g., the first detection of a binary black hole merger produced a strain of $10^{-21}$,\cite{GW150914} corresponding to a change in distance equal to a fraction of the width of a proton.
